---
title: "On Synergy"
author: "John Zobolas"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
description: "A book about synergy models"
url: 'https\://bblodfon.github.io/synergy-doc/'
github-repo: "bblodfon/synergy-doc"
bibliography: ["references.bib", "packages.bib"]
link-citations: true
site: bookdown::bookdown_site
---

```{r include = FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(.packages(), 'bookdown'), 'packages.bib')
```

# Preface {-}

Hi, there!
You might wonder why this notes/book?

In the beginning I made these notes because I found the need to clarify some things in my head regarding the mathematical models that are used in the world of drug combinations and their synergistic effects and how these models have been used to computationally assess synergy, antagonism or non-interaction between drugs that are tested in high-throughput screens. 
So, I read a *few papers* ^[Ok, maybe a lot of them!] and made a Word document with my notes.
Later I thought of making this as a bookdown [@R-bookdown] document since it would be nicier to have it in that format for online reading and surely much more *accessible* to the scientific community - and here we are.

Before you read further, let me assure you of this fact: **this is not a review paper or a book for the subject at hand**, more like nicely structured notes, and personal for that matter.

I have copied statements used in the papers exactly as they were written (so full plagiarism). 
They wrote it, they said it, it is what it is. 
I repeat: these are just **NOTES**. 
If these notes help you, then amazing!
If you want to contribute, you are most welcome!

And remember, as one of my supervisors once said:

> What is a mountain, what is a synergy?

...and after a while, we concluded at:

> The mountain *is* the synergy!

## Structure {-}

TODO

## Rsession-info {-}

The most important thing is to install bookdown [@R-bookdown]. 

```{r session-info}
xfun::session_info()
```